Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-29 manuscript/HIAs.qmd
fix methods map with infrastructure map

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# set eval to true to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

<!--# The document is set to eval=F in the yaml, and some code chunks are set manually to eval=T. When working on the manuscript, you cannot run all with  Ctr+Alt+R as this ignores the YAML. This will read inn data cashed locally. To also evaluate the more time consuming chunks, change the YAML to eval: true and run all (or render to html):  -->

@@

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)

@@

# Soem directories
mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

@@

path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
```

@@

# read cached rds file
naturetypes <- readRDS("../data/surveyData.RDS")
```

@@

table(nature3$municipality)

@@

# calculate stats - terrestrial area
terrestrial <- terrestrial |>
  mutate(
    area_t = geometry |> st_area(),
    t_area_km =
      round(units::drop_units(area_t * 1e-6))
  ) |>
  as_tibble()
```

@@

tmap_mode("plot")
positionMap <- tm_shape(outline) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    size = 1,
    xmod = 2,
    ymod = -1
  ) +
  tm_symbols(
    size = 2,
    shape = 17,
    col = "black"
  ) +
  tm_grid(projection = 4326) +
  tm_layout(bg.color = "skyblue")
```

@@

# tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_nf_stars)+
#  tm_raster(col="red")
# OK

@@

# Caululate the area of mire inside the coverage maps
# -- Nordre Follo
mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Nordre Follo")
```

@@

# path to local cahing folder
path_temp <- ifelse(dir == "C:",
  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/"
)

@@

# read cached vetorized infrastructure data
infra <- readRDS(paste0(path, "infrastructureIndex_discrete_vectorized.rds"))

@@

# Make a plot to check that it has worked
infraMuni3_tbl |>
  ggplot() +
  geom_bar(
    aes(
      x = infrastructureIndex,
      y = total_area_HIAs_km2,
      fill = factor(infrastructureIndex),
      colour = factor(infrastructureIndex)
    ),
    stat = "identity",
    lwd = 1.2
  ) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr")) +
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
  facet_grid(cols = vars(Municipality)) +
  theme_minimal_hgrid() +
  labs(
    x = "Human Impact Area (HIA) class",
    y = "Area (km<sup>2</sup>)"
  ) +
  theme(
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      orientation = "left-rotated",
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    strip.background = element_blank(),
    strip.text = element_textbox(
      size = 12,
      color = "white", fill = "#5D729D", box.color = "#4A618C",
      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
    )
  ) +
  guides(fill = "none", colour = "none")
```

@@

infraMuniMap <- tm_shape(muni3) +
  tm_borders() +
  tm_shape(infraMuni3) +
  tm_polygons(
    col = "infrastructureIndex",
    style = "cat",
    title = "Human Impact\nArea Class"
  ) +
  tm_shape(muni3) +
  tm_borders(lwd = 3, col = "black") +
  tm_facets(by = "Municipality") +
  tm_layout(legend.show = TRUE)

# Plot
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )

(methodsMap <- tmap_arrange(
  # positionMap,
  muniPlot,
  infraMuniMap,
  widths = c(.4, .6)
))

# cowplot::plot_grid(tmap_grob(muniPlot),
#                   tmap_grob(infraMuniMap),
#                   ncol=1,)

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

<!--# The document is set to eval=F in the yaml. This only affect the rendering, and not the interactive session. Eval is also set explicitly for each chunk. When eval is set to false it means the job is quite time consuming and I instead run it once and cache it locally.  -->

@@

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)
library(cowplot)

@@

# Some directories
mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

@@

path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

# path to local caching folder
path_temp <- ifelse(dir == "C:",
  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/"
)
```

@@

# read cached rds file. This is not so big.
naturetypes <- readRDS("../data/surveyData.RDS")
```

@@

nature3 |>
  as_tibble() |>
  count(municipality,
    sort = TRUE,
    name = "Number of polygons")

@@

# calculate stats - terrestrial area
terrestrial <- terrestrial |>
  mutate(
    area_t = geometry |> st_area(),
    t_area_km =
      round(units::drop_units(area_t * 1e-6))
  ) 
```

@@

(positionMap <- tm_shape(outline) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    size = 1,
    xmod = 2,
    ymod = -1
  ) +
  tm_symbols(
    size = 2,
    shape = 17,
    col = "black"
  ) +
  tm_grid(projection = 4326) +
  tm_layout(
    bg.color = "skyblue",
    outer.margins = c(0.01, .02, .02, .02)))
```

@@

@@

# Calculate the area of mire inside the coverage maps
# -- Nordre Follo
mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Nordre Follo")
```

@@

@@

# read cached vetorized infrastructure data
infra <- readRDS(paste0(path_temp, "infrastructureIndex_discrete_vectorized.rds"))

@@

# Make a plot to check that it has worked
infraMuni3_tbl |>
  ggplot() +
  geom_bar(
    aes(
      x = infrastructureIndex,
      y = total_area_HIAs_km2,
      fill = factor(infrastructureIndex),
      colour = factor(infrastructureIndex)
    ),
    stat = "identity",
    lwd = 1.2
  ) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr")) +
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
  theme_minimal_hgrid() +
  labs(
    x = "Human Impact Area (HIA) class",
    y = "Area (km<sup>2</sup>)"
  ) +
  theme(
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      orientation = "left-rotated",
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    strip.background = element_blank(),
    strip.text = element_textbox(
      size = 12,
      color = "white", fill = "#5D729D", box.color = "#4A618C",
      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
    )
  ) +
  guides(fill = "none", colour = "none") +
  scale_y_log10() +
  facet_grid(cols = vars(Municipality)) 
```

@@

# Infratructure in each municipality
infraMuniMap <- tm_shape(muni3) +
  tm_borders() +
  tm_shape(infraMuni3) +
  tm_polygons(
    col = "infrastructureIndex",
    style = "cat",
    title = "Human Impact\nArea Class"
  ) +
  tm_shape(muni3) +
  tm_borders(lwd = 3, col = "black") +
  tm_facets(by = "Municipality") +
  tm_layout(
    legend.show = FALSE,
    panel.label.height = 0)

# A figure with just the legend
infraMuniMap_l <- tm_shape(muni3) +
  tm_borders() +
  tm_shape(infraMuni3) +
  tm_polygons(
    col = "infrastructureIndex",
    style = "cat",
    title = "Human Impact\nArea Class"
  ) +
  tm_layout(legend.only = TRUE,
            legend.position = c("left", "bottom"),
            legend.outside = TRUE)

# Syrvey coverage and other colour overlayed the municipalities
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )

(methodsMap <- tmap_arrange(
  positionMap,
  muniPlot,
  infraMuniMap_l,
  infraMuniMap,
  widths = c(.35, .65),
  heights = c(.6, .4),
  outer.margins = NULL
))
